A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-06-10 ManuscriptGeophysics2018.tex
All answered; needs work on code/figs/manuscripts

All mentioned publications have in common that they were derived for direct
current methods (DC) or low frequency methods, such as time-domain shallow EM
methods (TEM), or controlled-source electromagnetics (CSEM), but not for high
frequency methods such as ground-pe\-ne\-tra\-ting radar (GPR). Generally it
was even thought that the filter method works only in the diffusive limit where
the quasistatic approximation is valid \citep[e.g.,][]{GEO.15.Hunziker}.
Nevertheless, \cite{GEO.17.Werthmuller} tested DLF for modeling 250 MHz
center-frequency GPR data using a 401 point filter that was designed for
diffusive EM, and found that it was orders of magnitude faster than quadrature
approaches. However, good quality results were only obtained for the first
arrivals at short time-offsets and the later arrivals had poor quality. Yet,
this promising initial test suggested it might be possible to specifically
design a special filter for GPR frequencies where the EM fields propagate as
waves rather than by diffusion.

@@

In the following sections we give a brief overview of the methodology, the
theory, and the code, and then show examples of its usage, as well as
presenting new filters for both CSEM and GPR data. The current article falls
therefore into all three mentioned categories by the new application of DLF to
wave phenomena and by providing new filters with the accompanying code. The
algorithm and many more examples of its usage can be found on
\href{https://github.com/empymod/article-fdesign}{github.com/empymod/article-fdesign}.
The examples can be used as templates to design new filters.

@@

In the direct matrix inversion method for solving for the digital filter
coefficients, equation~\ref{eq:filtapprox} is cast as a linear system\old{ as
follows.}\new{, where} $\Delta$ and $\delta$ are preassigned scalar values and
$r$ is a range of $M$ preassigned values $r_m$. The filter base coefficients
$b_n$ are computed using equation~\ref{eq:base} and an array of values for $l$
are computed using $l_{mn} = b_n/r_m$. The linear system's left hand side (LHS)
matrix $\bf{A}$ has dimensions $M \times N$ with coefficients \old{$A_{mn} =
f(l_{mn})/r_n$}\new{$A_{mn} = f(l_{mn})/r_m$} and the right hand side (RHS)
vector $\bf{v}$ has elements \old{$v_n = F(r_n)$}\new{$v_m = F(r_m)$}. The
resulting linear system
%
\begin{equation}
  \bf{A}\bf{h} = \bf{v}
  \label{eq:,linearSystem}
\end{equation}
%
can be solved by direct matrix inversion to obtain the vector of filter
coefficients $\bf{h}$. For a given filter length $N$, there are many subjective
choices that go into the practical implementation of this method, including the
choice of the transform pairs $F(r)$ and $f(l)$, as well as the value for $M$
and the particular range and spacing of the values $r_m$. Once values are
chosen for these variables, an optimal filter can be found by a grid search
over $\Delta$ and $\delta$ for values that produce a high quality filter. The
choice of metric for what constitutes a high quality filter is also subjective
and is further discussed below.

>>>>

All mentioned publications have in common that they were derived for direct
current methods (DC) or low frequency methods, such as time-domain shallow EM
methods (TEM), or controlled-source electromagnetics (CSEM), but not for high
frequency methods such as ground-pe\-ne\-tra\-ting radar (GPR). Generally it
was even thought that the filter method works only in the diffusive limit where
the quasistatic approximation is valid \citep[e.g.,][]{GEO.15.Hunziker}.
Nevertheless, \cite{GEO.17.Werthmuller} tested DLF for modeling 250 MHz
center-frequency GPR data using a 401 point filter that was designed for
diffusive EM, and found that it was orders of magnitude faster than quadrature
approaches. However, good quality results were only obtained for the first
arrivals at short time-offsets and the later arrivals had poor quality. Yet,
this promising initial test \old{suggested}\new{suggests} it might be possible
to specifically design a special filter for GPR frequencies where the EM fields
propagate as waves rather than by diffusion.

@@

In the following sections we give a brief overview of the methodology, the
theory, and the code, and then show examples of its usage, as well as
presenting new filters for both CSEM and GPR data. The current article falls
therefore into all three mentioned categories by the new application of DLF to
wave phenomena, \new{which means that the DLF can be used to model seismic
wavefields as well,} and by providing new filters with the accompanying code.
The algorithm and many more examples of its usage can be found on
\href{https://github.com/empymod/article-fdesign}{github.com/empymod/article-fdesign}.
The examples can be used as templates to design new filters.

@@

In the direct matrix inversion method for solving for the digital filter
coefficients, equation~\ref{eq:filtapprox} is cast as a linear system\old{ as
follows.}\new{, where} $\Delta$ and $\delta$ are preassigned scalar values and
$r$ is a range of $M$ preassigned values $r_m$. The filter base coefficients
$b_n$ are computed using equation~\ref{eq:base} and an array of values for $l$
are computed using $l_{mn} = b_n/r_m$. The linear system's left hand side (LHS)
matrix $\bf{A}$ has dimensions $M \times N$ with coefficients \old{$A_{mn} =
f(l_{mn})/r_n$}\new{$A_{mn} = f(l_{mn})/r_m$} and the right hand side (RHS)
vector $\bf{v}$ has elements \old{$v_n = F(r_n)$}\new{$v_m = F(r_m)$}. The
resulting linear system
%
\begin{equation}
  \bf{A}\bf{h} = \bf{v}
  \label{eq:,linearSystem}
\end{equation}
%
can be solved by direct matrix inversion\new{, or any other matrix inversion
routine,} to obtain the vector of filter coefficients $\bf{h}$. For a given
filter length $N$, there are many subjective choices that go into the practical
implementation of this method, including the choice of the transform pairs
$F(r)$ and $f(l)$, as well as the value for $M$ and the particular range and
spacing of the values $r_m$. Once values are chosen for these variables, an
optimal filter can be found by a grid search over $\Delta$ and $\delta$ for
values that produce a high quality filter. The choice of metric for what
constitutes a high quality filter is also subjective and is further discussed
below.
